# If / Else

Building on our logical operators, there will often be times where you want to split the logic of your code depending on a criteria. For example, if you've created a function that can accept a character string or a number, you might want to split the body of the function to do something slightly different depending on the class of the provided argument.

## Structure

If / else statements in R has a simple structure:
  ```{r, eval = FALSE}
if (criteria_statement) {
  what_you_want_to_do
} else if (other_criteria) {
  something_else_you_want_to_do
} else {
  something_you_want_to_do_if_all_else_fails
}
```

Putting this into practice, a real If / else block may look like this:
  
  ```{r}
x <- 1
if (x == 1) {
  return("x is 1")
} else if (x == 2) {
  return("x is 2")
} else {
  return("x is not 1 or 2")
}
```

Implementing this in a function could look like this:
  
  ```{r}
what_is_it <- function(x) {
  if (is.character(x)) {
    return("x is a character")
  } else if (is.numeric(x)) {
    return("x is numeric")
  } else {
    return("x is something else")
  }
}

what_is_it("hello")
what_is_it(2)
what_is_it(TRUE)
```

## Questions {#questions-if-else}

1. Rather than writing `if (x == 1 | x == 2 | x == 3)`, how could you use the `%in%` operator to make it shorter?
  2. How does the `switch()` function relate to if / else statements?