# Subsetting

There will be occasions where you don't want all the values in a vector/list/matrix/dataframe. Instead, you'll only want a *subset*. The way to do that is slightly different depending on the data structure you're using.

## Vectors

Vectors are simple. Just use square brackets (`[]` or `[[]]`) after your vector and provide the index or indices of the values that you want:

```{r}
c(10,20,30,40)[1]
c(10,20,30,40)[c(1,4)]
c(10,20,30,40)[1:3]

c(10,20,30,40)[[1]]
```

P.S. If you have a vector of named values, you can also use the names instead of the indices. Like `c(value_1 = 1)[["value_1"]]`.

But Adam, I hear you ask, `c(10,20,30,40)[1]` and `c(10,20,30,40)[[1]]` just gave us the same thing, so are the interchangeable?

Well, they kind of returned the same thing, but they didn't. So no, they're not interchangeable.

Essentially, `[]` returns the *container* at the provided index, where `[[]]` returns the *value* at the provided index. Let's see a practical example of how these are different:

```{r}
c(value_1 = 10,
  value_2 = 20)[1]
c(value_1 = 10,
  value_2 = 20)[[1]]
```

In the first call, we get the name of the value and the value itself. In other words, rather than just returning the value at that index, we've essentially just chopped up the vector to only returning everything from the first position. Conversely, in the second call, we've just been given the value. What we've done here is extracted the value out from that position.

/TO DO

I think some kind of analogy could be useful here to drive home the difference.

It's a very subtle difference, but it is an important one. Make sure that if you want the value, use `[[]]`, and if you want the whole part of the vector, use `[]`.

## Lists

Lists can be subsetted in the same way as vectors - `[]` returns the container at the index provided and `[[]]` returns the value:

```{r}
list(
  value_1 = c(1,2,3),
  value_2 = c("hello", "there", "everyone")
)[[1]]
list(
  value_1 = c(1,2,3),
  value_2 = c("hello", "there", "everyone")
)[1]
```

A key difference with lists however, is that you can also subset based on the name of the value in the list using the `$` operator:

```{r}
list(
  value_1 = c(1,2,3),
  value_2 = c("hello", "there", "everyone")
)$value_1
```

This is equivalent to:
```{r}
list(
  value_1 = c(1,2,3),
  value_2 = c("hello", "there", "everyone")
)[["value_1"]]
```

Another key difference is that lists can, of course, hold recursive values. This means that subsetting a list can return another list, that can also be subsetted and so on:

```{r}
list(
  list_1 = list(
    list_2 = list(
      list_3 = "hello"
    )
  )
)[1][1][1]
```

And of course, you can do the same thing with the `[[]]` operator if you only want the value and not the container.

## Matrices

Matrices are two dimension, meaning they can't be subsetted with a single value. Instead, we still use the `[]` operator, but we provide two values: one for the row and another for the column:

```{r}
matrix(c(1:10), nrow = 5, ncol = 2)
matrix(c(1:10), nrow = 5, ncol = 2)[4,1]
```

## Dataframes

Dataframes can be subsetted in the same way as matrices (using the `[]` operator). However, dataframes can also be subsetting (like lists), using the `$` operator and the name of the column:

```{r}
data.frame(
  col_1 = c(1,2,3),
  col_2 = c("hello", "there", "everybody")
)$col_1
```

Why does this approach work for dataframes? Well, as I alluded to [before](#dataframes), dataframes store columns as lists. But technically, the dataframe itself is also stored as a kind of list, with each column being another entry in that list. So, just like we can subset lists using `$`, we can subset dataframes with it as well because a dataframe is like a fancy list.
