# For Teachers

Now we have a basic understanding of some of the core concepts in R, let's have a look at them in a bit more detail. In contrast to the previous section, here were going to focus on the underpinnings of R. You can kind of think of this section as being less practical and more theoretical.

If you're going to teach R, this section will provide you with a good level of understanding. Hopefully, you'll not only know *how* to do something when asked, but *why* you should do it that way.

## Functions

In the "For Students" section, we looked at what a function is and how to use one. In this section, we're going to look more at the structure of a function and how you might go about writing your own functions.

### Creating functions

R and its packages give you access to hundreds of thousands of different functions, all tailored to perform a particular task. Despite this wide array to choose from however, they will always be cases where there isn't a function to do exactly what you need to do. For those of you coming over from Excel, this can often be a serious source of frustration where there isn't an Excel function for you to use and there isn't an easy way to create one without knowing VBA.

R is different. Creating functions can be very simple and will really change the way you work.

Creating functions will also highlight an important delineation. Previously, we've been focusing on *calling* functions. Calling a function is essentially using it. But in order to call a function, it needs to be *defined*. Base functions are already defined (i.e. someone has already written what the function is going to do), but when you're creating your own functions, you are *defining* a new function that you're presumably going to call later on.

#### Function structure

If we go back to the beginning of this chapter, we learned that everything that exists is an object. Functions are no exception, and so we create them like we do all our other objects. There is a slight diversion however. When we define a function, we assign it to an object with the `function` keyword like this:

```{r}
my_first_function <- function() {}
```

Notice how we've got two sets of brackets here. The first (`()`) is where we define our input parameters. The second (`{}`) is where we define the body of our function.

Let's do a simple example. Let's create a function that adds two numbers together:

```{r}
my_sum_function <- function(x, y) {
  x + y
}
```

So in this example, I've defined that when anyone uses the function, they need to provide two input parameters named `x` and `y`. Something that people tend to struggle with is that the names of your input parameters have no implicit meaning. They are just used to reference the value provided in the body of the function and, hopefully, make it clear what kind of thing the user of the function should be providing. This is why for example in some functions that require a dataframe there will be an input parameter called `df` or similar. But importantly, these names are technically just arbitrary.

In the body of the function, we can see that we're just doing something really simple: we're adding `x` and `y` together with `+`.

Once I've run the code to **define** my function, I can then **call** it like I would any other function:

```{r}
my_sum_function(x = 5, y = 6)
```


##### Optional input parameters

When defining your function, you can define optional parameters. These will likely be values where most of the time you need it to be one thing, but there are edge cases where you need it to be something else. Defining optional parameters is really easy; whenever you define your function, just give it a value and that will be its default:

```{r}
add_mostly_2 <- function(x, y = 2){
  x + y
}

add_mostly_2(x = 5)

add_mostly_2(x = 5, y = 3)
```

#### ...

You'll notice a crucial distinction between R's `sum()` function and ours. The base function allows for an indeterminate number of input parameters, whereas we've only allowed 2 (`x` and `y`). This is because the base `sum()` function uses a `...`. This `...` is essentially shorthand for "as many or as few inputs as the user wants to provide". To use the `...`, just add it as in an input parameter:
```{r}
dot_dot_dot_function <- function(x, y, ...) {
}
```

The `...` works particularly well when you might be creating a function that *wraps* around another one. A wrapping function is just a function that makes a call to another one within it, like this:

```{r}
sum_and_add_2 <- function(...){
  sum(...) + 2
}
```

All we're basically doing in the above wrapping around the `sum()` function to add some specific functionality.

By using the `...` here, we can just pass everything that the user provides to the `sum()` function. This means we don't have to worry about copying any input parameters.

##### Return values

As I mentioned in the "For Students" section, functions have a single return value. By default, a function will return the last evaluated object in the function environment. In our `my_sum_function` example, our last evaluation was `x + y`, so the output of that was what was returned by the function.

You can also be explicit with your return values by using the `return()` function. The `return()` function will return whatever is provided to the `return()` function. This can be useful if you want to return a value prematurely:

```{r}
early_return_function <- function(x,y, return_x  = TRUE) {
  if (return_x) {
    return(x)
  }
  x + y
}
```

Here, we can see more clearly that `x` is returned when `return_x` is TRUE and `x + y` is returned otherwise.

Certain style guides suggest that you should **only** use `return()` statements for early returns. In other words, the "normal" return value for your function should be defined by what's evaluated last. Personally, I think you should use whatever makes it clearer for you. I quite like seeing explicit `return()` values in a function because I find it makes it clearer what all the possible return values are, but this is just personal preference.

If you're interested, there are some specifics about function inputs in R that can be good to know.

### Input validation

Firstly, unlike some other languages, functions do not have a specific data type tied to each input parameter. Any requirements that should be imposed on an input parameter (e.g. it should be numeric) are done by the function creator in the body of the function. So for instance, when you try to sum character strings, the error you get occurs because of type-checking in the body of the function, not when you provide the input parameters.

EXAMPLES


### Functions as objects

Secondly, functions are technically just another object. This means that you can use functions like you would any other object. For instance, some functions will accept other functions as an input parameter. When we move onto the `apply` logic, the `lapply()` (list-apply) function requires a `FUN` parameter that is the function the be applied to each value in the provided list.

```{r}
sum_list <- list(
  c(1,2),
  c(5,10),
  c(20,30)
)

lapply(sum_list, FUN = sum)
```

Linked with the idea that functions are just another type of an object, there is an important distinction between 
`sum`
and
`sum()`. The first will return the `sum` *object*. That is, not the result of applying inputs to the `sum` function, but the function itself. If you just type the name of the function into the console, it will show you the code for that function (it's definition): 

```{r}
sum
```

Conversely, `sum()` will *call* the `sum` function with the inputs provided in the brackets.

EXAMPLE



## Environments 

TO DO

### Environment basics

At its core, an environment is a collection of objects. A bit like a list, environments store multiple objects in a single structure.

To create a new environment, we use the `new.env()` function.

```{r}
new_env <- new.env()
```

To add items to our environment, we can add them like we would a list using the `$` operator:

```{r}
new_env$first_object <- "hello"
```

To list all of the objects in an environment, we use the `ls()` function:

```{r}
ls(new_env)
```

Importantly, you can't have two objects in the same environment with the same name. If you try, you'll just overwrite the previous value:

```{r, error = TRUE}
new_env$first_object <- "world"
new_env$first_object
```
### Environment inheritance

Environments have parents and children. In other words, there is a hierarchy of environments, with environments being encapsulated in other environments while also encapsulated other environments.

Every environment (with the exception of what we call the empty environment) has a parent. For example, when I created my `new_env` environment before, this was created in the **global environment**. The global environment is the environment that objects are assigned to when working in R interactively. The global environment's parent environment will be the environment of the last package you loaded. Packages have environments to avoid name conflicts with functions and to help R know where to look for a function.

At the top of the environments of packages you've loaded will be the **base environment** which is the environment of base R. Finally, the base environment's parent is the **empty environment** which does not have a parent.


The hierarchy of these environments looks like this:

```{r, echo=FALSE}
DiagrammeR::grViz("
digraph environments {

  # a 'graph' statement
  graph [overlap = true]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica,
        fillcolor = Grey,
        style= filled]
  'empty environment';
  'base environment';
  'package environment\n(e.g. package:BMRSr)';
  'global environment';

  # several 'edge' statements
  'empty environment'->'base environment'
  'base environment'->'package environment\n(e.g. package:BMRSr)'
  'package environment\n(e.g. package:BMRSr)'->'global environment' 
}
")
```

### Scope

So we know that objects in the same environment can't have the same name, but what happens when two different environments happen to have objects with the same name? This is where the concept of **scope** comes in. **Scoping** is the set of rules that governs where R will look for a value.

Well, R will search for the object in order of environment, starting at the most specific environment (so the global environment in the above diagram) and moving up. For example, we know that there is a function in base R called `sum()`. But, if I define a new function in the global environment called `sum` then which function will be called when I type `sum(...)`. Well, because we know that the search path starts from the most specific environment, R will look for `sum` in the global environment first and it'll find the `sum` that I've just defined. At this point, it'll stop looking because `sum` has been found.

For this reason, it's a good idea to use a package like `conflicts` to  manage the packages you loaded, otherwise which function you use when you have two functions with the same name from different packages will be defined by which one you loaded later.


### Function environments

Functions, when they called, create their own more specific environment. The parent of this environment will be the environment in which it was called (most often this will be the global environment).

This breeds some specific behaviours. For example, say you've written a function that expects two input parameters, `x` and `y`. Well, what would happen if someone had already defined an `x` and `y` variable in their script? Which value should R use?

Let's see what happens.

```{r}
sum_custom <- function(x,y) {
  x + y
}

x <- 10
y <- 5

sum_custom(x = 1, y = 2)
```

In this case, the fact that there is already an `x` and `y` in the global environment doesn't really make much difference. The function creates its own more specific environment when it's called, and it looks for the `x` and `y` variables in here first. It finds them and uses those values (1 and 2).

But what happens if a variable doesn't exist in the more specific function environment? Let's take a look.


```{r}
sum_custom <- function(x,y) {
  x + y + w
}

w <- 5

sum_custom(x = 1, y = 2)
```

In this case, the function looks in the specific environment for `w`, but it doesn't exist. The only objects that exist in the function environment are the `x` and `y` that we've provided. So when R doesn't find it in the more specific environment, it looks in the less-specific global environment. It finds, and so it uses the value it finds.

This can be a dangerous thing, so always make sure that you're function is accessing the values you think it is.

So does R work the other way? Does it ever look in a more specific environment? Nope. 

```{r, error = TRUE}
sum_custom <- function(x,y){
  im_a_sneaky_variable <- 10
  x + y
}

im_a_sneaky_variable
```

Once the function is called, objects in its environment are inaccessible. The long and short of it is, R will start from specific environments and then look upwards, never downwards.


#### 'Super' assignment

There will be occasions however when you need to make changes to the global environment. For instance, say you want to increment a counter every time a function is called, regardless of where it's called from. In these cases, you can use the controversial `<<-` operator. This is used as an assignment operator to assign a value to the global environment. Observe...


```{r}
sum_custom <- function(x,y) {
  count <<- count + 1
  x + y
}

count <- 0

sum_custom(1,2)

count

sum_custom(2,3)

count
```
Note how when we assign `0` to our `count` variable outside of the function, we don't need to use `<<-`. This is because we're already assigning to the global environment.

Use the `<<-` with care and only assign something to the global environment if you really need to. Otherwise, you may start overwriting variables in your global environment without ever realising it.


## Objects and Classes

Objects and classes are a fairly ubiquitous concept across programming languages and data analysis tools. We'll briefly look at what an object is in R, but by no means is this an exhaustive description.

### What is a class?

The world is complicated. Everything in the world is unique and defined by an infinite number of properties and features. For example, if you take a person, they can be defined by their name, where they're from, where their parents are from, their hair colour, their likes and dislikes, and so on and so on. When we want to store data in a structured and formal way in a computer system, however, this isn't particularly helpful. Instead, we need to store data in a predefined structure. This is essentially what an class is; a pre-defined structure to store important attributes or features of a **thing**.

Let's take the person instance again. Let's say we're going to store data on a number of individuals, we won't be able to store everything about them. So we'll choose a subset of their attributes or features to store that are relevant to what we need. But to make things more efficient, we're going to store the same information for each one. So lets say we're going to do some geographical analysis, we might want to include a person's name, their nationality, and perhaps their ethnicity. So for each person we want to store, we can store these three attributes. And we might call this data structure a "person". Well this is exactly what an class is; a collection of attributes and features that is shared objects instances of the same type.

So our class is "person", and it has the attributes "name", "nationality", and "ethnicity". Now this obviously doesn't capture everything about a person, but it's enough for what we want to do.

Graphing that object might look something like this:

```{r echo=FALSE}
DiagrammeR::grViz("
digraph class {

  # a 'graph' statement
  graph [overlap = true]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica,
        style = filled,
        fillcolor = Grey]
  Person

  node [shape = circle,
        fixedsize = true,
        width = 0.8,
        fontsize = 8] // sets as circles
  Name; Nationality; Ethnicity

  # several 'edge' statements
  Person->Name Person->Nationality Person->Ethnicity 
}
")
```


### What is an object?

An object is just an instance of a class. So if you create a person from our "person" called John, then "John" is the object an "person" is the class.

When you create a dataframe for example, you're creating a `data.frame` object from the basic structure of the `data.frame` class.

### Objects and Classes in R

Looking more specifically at R, what kind of objects do we see. Well, according to John Chambers, the founder of the S programming language upon which R is based, everything is:

> "To understand computations in R, two slogans are helpful:   
>
>  *  Everything that exists is an object.
>  *  Everything that happens is a function call.
>
> --- John Chambers"

So every function, dataframe, plot, list, vector, integer, everything is an object.

To see the class of an object in R, use the `class()` function.

```{r}
class(data.frame()) # here we're finding the class of an empty data.frame
class(data.frame) # here we're finding the class of the data.frame() function
class(1)
```

### Inheritance

Sometimes, different classes are interrelated. For example, if you were storing data on vehicle, then you might have a "vehicle" class. But if you're storing data on lots of different vehicles, you might also have classes for each type of vehicle (e.g. "car", "motorbike", etc.). All of these classes are still vehicles, and so you don't want to have to repeat yourself when you define each of those classes. In other words, all of those classes are going to have some common attributes like colour, horsepower and so on. Similarly, each different type of vehicle will have some attributes that are unique to that type of vehicle. For instance, motorbikes can have sidecars but vans and cars don't. Cars and vans have doors but motorbikes don't. This highlights the benefits of inheritance. By creating a "vehicle" class and allowing your subsequent classes to inherit all of the attributes of the "vehicle" class, you can avoid duplication while allowing distinct classes. This way, when you want to add any attributes to all of your vehicles, you can just do it via the "vehicle" class rather than changing each type of vehicle class.

Diagramming this relationship:

```{r echo=FALSE}
DiagrammeR::grViz("
digraph inheritance {

  # a 'graph' statement
  graph [overlap = true]

  # several 'node' statements
  node [shape = box,
        fontname = Helvetica,
        style = filled,
        fillcolor = Grey]
  Vehicle;

  node [shape = box,
        fixedsize = true,
        width = 0.6,
        fontsize = 7,
        style = filled,
        fillcolor = Grey] // sets as circles
  Car; Motorbike; Van
  
  Vehicle->Car
  Vehicle->Van
  Vehicle->Motorbike

}")
```

Inheritance is an extremely deep topic which we won't go into here, but R objects also use inheritance. To see the inheritance tree of an object, use the `is()` function:

```{r}
is(1L)
```

Here we can see that an integer object is made up of the integer, double, numeric classes. The order of inheritance goes from left to right so we know that the integer class inherits from the double class, which inherits from the numeric class.

#### Object-Oriented Systems

Unfortunately (or fortunately, depending on your point of view), R doesn't have a single way of storing objects. In fact, there are 2 object-oriented (OO) programming systems in base R, and more (like R6) can be added via packages. These two base OO systems are S3 and S4, and they differ in the way that objects are constructed, stored, and interacted with. We're not going to go into the difference here, but I recommend Hadley's [Advanced R](https://adv-r.hadley.nz/) to better understand the difference between the two. For now, I'm just going to explain the basics of the S3 system as I think it's the easiest to understand and helps convey the philosophy behind why we use objects slightly more easily.

##### S3

In the S3 system, we rely on generic functions and methods. Generics are functions that have a single common goal, but that can be used for objects that might be very different. For example, `print()`-ing a dataframe is going to be different to `print()`-ing a plot or an API response or whatever. But `print()` always seems to know what to do. The reason it does is that the `print()` function is a **generic** function that actually uses a more specific function to achieve it's goal. In other words, we achieve a fairly high level goal like printing by calling a function that is specific to the object we're working on under the hood. These more specific functions are called **methods**.

As a real world analogy, think of the process of talking to someone. The common goal in talking is to communicate. But, depending on the language that someone speaks, the actual act of talking is going to be slightly different for different people you talk to. In this case, you can think of **communicating** as being the **generic** - it's the eventual goal. And **talking in the appropriate language** as being the **method**.

Going back to R, if you type `print.` into the console, the autocompleter will show you lots and lots of `print.something()` functions. These are all of the **methods** for all of the different printable objects in R. `print.date()` will print a date object, `print.data.frame()` will print a dataframe object and so on. But when you just use the `print()` function on an object, R will automatically choose which method it needs for the object you've passed as an input parameter.

```{r}
# here we're using the generic
print(as.Date("2020/06/10")) 
# because we're printing a Date object,
# this is the method that is actually used
print.Date(as.Date("2020/06/10"))
```

While you can often tell if something is an S3 method by it being a generic followed by a `.` and then an object name, don't rely on this, because people often use `.` to separate words in functions that can make them look like S3 methods when they're not. 

The important thing I hoped you've gleaned from this explanation is that you don't really need to know what your object is to use a generic. Yes, there are generics that only have methods for specific objects and so it may be important to know what type your object is to make sure you can use a particular generic, R takes care of which method is dispatched for you. If that doesn't convince you that you don't need to understand the underlying structure of an object to get hands on with R, I don't know what will.

Note: In some languages, methods are essentially functions or sub-routines that are tied to an class. For instance, a class that represents a person's bank account might have the method `Balance()`, which will return how much money a person has in their account. There are certainly some similarities between how methods are used in R and some other languages, they are a bit different. Mainly, in R, methods are not attributes of a class but are separate functions.

#### Creating objects

The obvious question is "Can I make my own objects in R?". The answer is yes. And quite easily too. Whether you should or not is a different story. I'm not going to go over how to create an object in S3 or S4 or any other OO system here because there is more than enough for an entire book, but there are plenty of resources out there that can help you.
