---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```

## Module 1 (Introduction to R)

Below are some interactive exercises to test your R knowledge. Just type your answers into the appropriate box, and hit "Run". In some cases, there'll be a hint or solution for you to use, but usually there'll be more than one way to do it!

### Exercise 1 (R and RStudio):
```{r r-rstudio, echo = FALSE}
quiz(
  question("R is best described as a...",
           answer("Object-oriented programming language."),
           answer("General purpose programming language."),
           answer("Statistical programming langauge.", correct = TRUE)
  ),
  question("RStudio is an integrated development environment (IDE) for R, which can be used to make interacting with the R language easier.
           Which of these statements about RStudio is *NOT* true?",
           answer("RStudio is the only IDE for R.", correct = TRUE),
           answer("RStudio sends the code you type to R to be executed."),
           answer("RStudio has many different tabs that show different things such as plots or files or scripts."),
           answer("You can't use R without RStudio.", correct = TRUE),
           incorrect = "Incorrect. There are many IDEs for R, and R can be used via its GUI without RStudio or any other IDE."
  )
)
```

### Exercise 2 (Arithmetic Operators):

Write the R code required to add 2 plus 2:

```{r two-plus-two, exercise=TRUE}

```

```{r two-plus-two-solution}
2 + 2
```


Write the R code required to divide 10 by 5:

```{r ten-divide-five, exercise=TRUE}

```

```{r ten-divide-five-solution}
10 / 5
```


### Exercise 3 (Variables)

Create a variable called `variable_1` and assign it the value 10. Then output it to the console.

```{r variables-ten, exercise = TRUE}

```

```{r variables-ten-solution}
variable_1 <- 10
variable_1
```

Create two numeric variables (called whatever you like), and multiply them.

```{r variables-multiply, exercise = TRUE}

```

```{r variables-multiply-solution}
variable_1 <- 10
variable_2 <- 5
variable_1 * variable_2
```


### Exercise 4 (Data Structures)

Create a *named* vector with 3 values. (Hint: That doesn't mean create a variable, it means that the values in the vector should be named)

```{r vector-named, exercise = TRUE}

```

```{r vector-named-solution}
c("value_1" = 10, "value_2" = 20, "value_3" = 30)
```

Create a list with two values; one that's a vector and another that's a dataframe.

```{r list, exercise = TRUE}

```

```{r list-solution}
list(
  "vector" = c(1,2,3,4),
  "dataframe" = data.frame(col_1 = "hello",
                           col_2 = "world")
)
```

### Exercise 5 (Subsetting)

Change the following code to only return the first 10 values from the vector.

```{r, subset-vector, exercise = TRUE}
vector_1 <- c(56,3,324,6,7,86,5,4,17,54,8468,45,345,234,7,56,8,4,32,234,5,764,658,6,3463)
vector_1
```

```{r, subset-vector-solution}
vector_1 <- c(56,3,324,6,7,86,5,4,17,54,8468,45,345,234,7,56,8,4,32,234,5,764,658,6,3463)
vector_1[1:10]
```

## Module 2 (Data Analysis)


