---
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    css: ["../../css/custom.css"]
  pdf_document: default
  md_document: default

---

class: center, middle

# Introduction to R

Adam Rawles
---

## Overview

--

- What is R and RStudio?

--

- Basic arithmetic operators

--

- Variable assigment

--

- Data types

--

- Data structures (including subsetting)

--

- Functions

---

## Learning Objectives


- Understand the basics of RStudio

--

- Know the difference between the data types and structures in R

--

- Understand the concepts of how functions operate in R

--

- Know where/how to look things up!

---

## What is R?


- Background

--

   + Public license programming language
   
--
   
   + Used mostly for statistical computing
   
--

   + Supported with many packages

--

- Interface

--

   + R uses a command-line interface

--

   + But there are many GUIs available (such as RStudio)

---

## RStudio

- RStudio is one of the most popular R environments

--

- RStudio has 4 main panes:

---

## RStudio

- Source

--

   + This is where your scripts (pre-written lines of codes) are displayed

--

   + If you click on a variable in the Environment pane, this will also be displayed in the source window

--

- Console

--

   + This is where the commands are actually executed
   
--

   + When you run a script from the source, it is passed to the console line by line

---

## RStudio

- Environment

--

   + This pane displays any variables or data structures you've created

--

   + This pane will also display any custom functions you've created, which we'll move onto in later modules
   
--

- History/Files/Plots/Help

--

   + This pane can display a number of things

--

   + The most useful of which are the Files tab, which displays all the files in the folder you're working in and the Plots tab, which will show any plots that you've created


--


   <b>Note</b>: You can move which pane shows which windows via the Tools -> Global Options dialog

---

## Basic arithmetic

- The basic arithmetic operators in R are very similar to Excel

--

```{r basic_arithmetic, echo = TRUE}
2 + 2 #add

2 * 3 #multiply

```

--

```{r basic_arithmetic_2, echo = TRUE}

6/2 #divide

10 - 5 #subtract
```

---

## Logical operators

- Logical operators in R are a bit different

--

```{r logical_operators_1, echo=TRUE}

1 == 1 #equal

1 != 2 #not equal

```

---

## Logical operators

```{r logical_operators_2, echo = TRUE}

2 > 1 #greater than

1 < 2 #less than

```

```{r}

1 == 1 | 1 == 2 #or

1 == 1 & 1 == 2 #and

```

---


## Variable assignment


- Assigning variables allows you store values for later use

--

```{r variable_assignment_1, echo = TRUE}
variable_1 <- 5
variable_2 <- 10

variable_3 <- variable_2/variable_1
variable_3

```
--

- For variable assignment, the `<-` and `=` operators can be used interchangeably
- But it's better to use `<-`

--

- <b>Note</b>: the value of a variable is not printed when it is assigned (i.e. when we assign `5` to `variable_1`, we don't get any output in the console)

---

## Variable assigment


- Variables can also be modified after they've been assigned

--

```{r variable_assignment_2, echo = TRUE}
variable_1 <- 1
variable_1

variable_1 <- "hello world"
variable_1

```

---

## Data types

- Data comes in lots of different forms

--

- Is "True" equal to TRUE?

--

- The main data types are:

--

+ logical; TRUE, FALSE

--

+ double; 12.5, 1, 999

--

+ integer; 2L, 34L, 1294L

--

+ character; "hello world", "True"

--

+ dates; 2019-06-01

--

+ datetimes; 2019-06-01 12:00:00

---

## Data types

- Assigning the right data type is important, as it determines how the data is stored and how data can be manipulated
--

```{r data_types_1, echo = TRUE, error= TRUE}

variable_1 <- 5
variable_2 <- "5"

variable_1 + variable_2
```

---

## Data types

- R will automatically decide the data type when you assign a variable, but you can force R to store the value as a different data type using the as.xxxxx functions
--

```{r data_types_2, echo = TRUE}
variable_1 <- as.numeric("5")
variable_2 <- as.integer("5")

variable_1 + variable_2

```
--

- But the value has to be coercible in the first place!

--

```{r data_types_3, echo = TRUE, error = TRUE}
variable_1 <- as.numeric("hello world")
variable_1
```

---

## Data types - Factors

- There is another data type which is fairly common in R called a factor

--

- A factor is made up of one or more levels that represent some form of grouping

--

- For example, if we had a dataframe containing firms from different division, we might store "Lead Division" as a factor, with the levels Banking, Insurance, Investment, Fiduciary

--

- <b>Note</b>: when importing data into R, R will automatically try and convert columns containing strings into factors (unless you specify otherwise)

---

## Data structures

- Values in R need to be stored in a specific way
- There are 4 main data structures in R:

---

## Data structures - Vectors

- Vectors

--

  + Vectors in R are arrays of data in one dimension

--

  + They are atomic (not recursive)*, but can have named values

--

  + Even variables with a single value will be stored as a vector with length 1

--


```{r data_structures_vectors, echo = TRUE}

vector_1 <- c(1,2,3,4,5,6)
vector_1

vector_2 <- c("first_value" = 1)
vector_2
```

- *More on the difference between atomic and recursive later
---

## Data structures - Matrices

- Matrices

--

  + Matrices are two-dimensional arrays that hold values of the same type
  
--

  + Matrices can have named rows or columns, but they cannot be subsetted by those names (more later on)
  
--

```{r data_structures_matrices, echo = TRUE}

matrix_1 <- matrix(c(1,2, 3,4), nrow = 2, ncol = 2)
print(matrix_1)

```
---

## Data structures - Data frames

- Data frames
  +  Most data you'll be working with in R will be held in a data frame
  
--
  
  +  Data frames are two dimensional arrays that can hold values of different types and have named columns and rows

--

```{r data_structures_dataframes, echo = TRUE}
dataframe1 <- data.frame(numeric_col = c(1,2),
                         character_col = c("hello", "world"),
                         stringsAsFactors = FALSE)
dataframe1
```

- Important notes:

--

  + Values in the same column must be of the same type

--

  + Each column must have the same number of rows

---

## Data structures - Lists

- Lists
  + Lists are similar to vectors in that they can contain multiple named values

--

  + However, lists can contain values of any type, including other lists

--

  + This means you can have a list of data frames, or a list of lists of data frames!
  
--

```{r data_structures_lists_1, echo = TRUE}

list_1 <- list(numbers = c(1,2,3,4),
               characters = (c("hello", "world"))
)
list_1

```

---

## Data structures - Lists

- Due to this difference between lists and vectors, we refer to a list as being *recursive* and vectors as *atomic*

--

- Atomic
  + "of or forming a single irreducible unit or component in a larger system."
    
--

  + Cannot hold objects of their own type (i.e. you can't have a vector object in a vector)
  
--
  
- Recursive
  + "characterized by recurrence or repetition."
  
--

  + Can hold objects of their own type (i.e you can have a list, in a list, in a list...)

---

## Subsetting data structures

- More often than not, you'll only want to access one or more of the values in a data structure

--

- For example, you may only want to see the first 10 values of a very long vector

--

- For vectors, this is done by suffixing the variable with a [] containing the index or indices of values you want

```{r subsetting_data_structures, echo = TRUE}
values <- c(6,45,7,54,99,31,12,15,67,100)
values[1]
```

---

## Subsetting data structures


```{r subsetting_data_structures_2, echo = TRUE}
values[c(1,10)]
values[1:5] # this is the same as values[c(1,2,3,4,5)]

```

---

## Subsetting data structures

- In R, you can also subset most data structures using [[]] instead of [], but there are two important differences

--


- [] returns the *container* of the value, not the value

--

- This will include the name of the value

--

- In a list, this will return a named list, rather than a single value (an example to follow)

--

- [] can be used with more than one index (e.g. `[c(1,2,3,4,5)]`)

--

- [[]] returns the *value* at that index, not the container

--

- [[]] will only accept a single index

---

## Subsetting vectors - exercise

- Create a vector

--

- Return the 5th value

--

- Return the 1st and 3rd value

--

- Return the 1st, 2nd, 3rd, 4th, 5th and 7th value

--

```{r subsetting_vectors_exercise}
vector1 <- c(1,5,8,9,67,5,3,2,5,6)
vector1[[5]]
vector1[c(1,3)]
vector1[c(1:5, 7)]
```

---

## Subsetting matrices


- For data structures with more that one-dimension, we have to provide some more information

--

- In these cases, we specify an index for each dimension, starting with the row

```{r subsetting_matrices, echo = TRUE}

matrix_1 <- matrix(c(1,2,3,4), ncol = 2, nrow = 2)
matrix_1[1,2]
matrix_1[,2]
```

--

- Matrices can also be subsetted by providing the names of the row/column instead of the number, or by using [[]]

---

## Subsetting data frames

- Unlike matrices, data frame columns can be referred to be name using the $ operator

--

- You can combine the $ and [] operators to access specific values in a column

--

```{r subsetting_data_frames, echo = TRUE}
dataframe_1 <- data.frame(numeric_col = c(1,2), character_col = c("hello", "world"), stringsAsFactors = FALSE)
dataframe_1$character_col
dataframe_1$character_col[1]
```

--

- <b>Note</b>: data frames can also be subsetted using the [x, y] format, with names or indices

---

## Subsetting dataframes - exercise


- Create a data frame with 2 columns

--

- Return one column using the $ operator

--

- Return the other column using the [,] approach

--

```{r subsetting_exercise, echo = TRUE}
df1 <- data.frame(numeric_col = c(1,2,3), logical_col = c(TRUE, FALSE, TRUE), stringsAsFactors = FALSE)
df1$numeric_col[1]
df1[,2]

```

---

## Subsetting lists


- Lists can also be subsetted using the $ and [] operators

--

```{r subsetting_lists_1, echo = TRUE}
list_1 <- list(numbers = c(1,2,3,4), characters = (c("hello", "world")))

list_1$numbers
list_1$characters[1]
```

--

- NB: the `$` operator cannot be used on atomic data structures (like vectors or matrices)

---

## Subsetting lists

- Although lists are technically two-dimensional, they do not have rows and columns and so cannot be subsetted using the [x, y] format

--

- Instead, lists are subsetted using the [] and [[]] operators

--

- These operators can then be combined with whatever operator is appropriate for the type of structure in the list (e.g. $ for a data frame, or [] for a vector)

--

```{r subsetting_lists_2, echo = TRUE}
list_1 <- list(numbers = c(1,2,3,4), characters = (c("hello", "world")))
list_1[[1]][2]

```

---

## Functions

- Any operation, such as loading a dataset or finding the mean, are defined as functions in R

--

- A function is a set of steps that takes an input, does some form of computation, and returns an output

--

- For example, the mean function takes values and returns the mean of those values

--

```{r functions_1, echo = TRUE}
values <- c(10,20,30)
mean(values)

```

--

- You may have noticed that `c()` also acts as a function, creating a vector with the provided values

---

## Functions - input parameters/arguments

- Almost all functions will require some input to produce an output

--

- These are the function's input parameters or arguments

--

- For example, in the previous slide, the mean() function required a vector of values to calculate the mean

---

## Functions - input parameters/arguments

- Some functions will have optional input parameters

--

- To see what arguments/parameters a function requires, type the function name preceded by a "?" in the R console (e.g. `?mean`)

---

## Functions - exercise

- Find the help page for the `sample()` function

--

- Use the `sample()` function

---

## Functions - naming input parameters

- When you provide unnamed values to a function, R will automatically assign them to an input argument in the order they appear on the function's help page

--

- This can be hard to keep track of if you're providing multiple arguments

--

- Best practice, therefore, is to be explicit when providing your input arguments

--

- Let's use the `substr()` function as an example

---

## Functions - naming input parameters


- The `substr()` function returns part of a string from a start and stop index

--

```{r function_naming_input_parameters_1, echo = TRUE}
string1 <- "hello world"
substr(string1, 1,5)
```

- The `substr()` function expects 3 arguments;

--

- The string (x)

--

- The start point (start)

--

- The end point (stop)

--

- As shown above, we can just pass those arguments without naming them, in that order

---

## Functions - naming input parameters

- It's best practice, however, to name the arguments as you pass them...

--

```{r function_naming_input_parameters_2, echo = TRUE}
string1 <- "hello world"
substr(x = string1, start = 1, stop = 5)
```

---

## Recap

- R is a free statistical programming language
- One of the most popular environments for R is RStudio
- Mathematical operators in R are mostly the same as in Excel (e.g. +, -, *, /...)
- Values in R can take different forms, and it's important to get the form correct!
- Values can then be stored in lots of different structures
- These structures can be subsetted using different approaches ([], [x,y], $, [[]])
- Almost every action in R is performed via a function
- A function takes an input, does some computation and returns an output
- It's always good to be explicit when providing arguments to functions
