---
title: "R Cheat Sheet"
author: "Adam Rawles"
output: html_document
css: cheat_sheet.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```


<div class="col_3_out">
<div class ="col1">


###For loops

Use for loops to do something with every value in a vector/list or to do something a certain number of times

```{r for_loops, eval = FALSE}
for (identifier in vector/list){
  do something
}

for (identifier in 1:somenumber){
  do something
}
```


<hr></hr>
###If else statements

Use if else statements to only do something if one criteria (or more) is fulfilled

```{r if_else, eval = FALSE}
if (value1 == value2){
  do this
}
else if (value1 < value3){
  do this
}
else {
  do something else
}
```
</div>
<div class = "col2side2">


###Connecting to SQL Databases

```{r sql, eval = FALSE}
library(obdc)
con <- dbConnect(odbc(), Driver = "SQL Server", Server = e.g "PSADS004", Database = e.g. "TrailBlazer", trusted_connection = TRUE)
```


<hr></hr>

###Querying SQL Databases
Note: you must connect to a database first

```{r querying, eval = FALSE}
result <- dbGetQuery(conn = con, statement = "SELECT * FROM table1...")
```


<hr></hr>
###Summary statistics

```{r summary_statistics, eval = FALSE}
mean(df$column) #Returns the mean
median(df$column) #Returns the median
sd(df$column) #Returns the standard deviation
quantile(df$column) #Returns the 0, 25th, 50th, 100th quantiles
```

</div>
<div class = "col3side2">

```{r summary_statistics_2, eval = FALSE}
summary(df) #Returns summary stats for all
#columns in df
```

```{r summary_statistics_3, eval = FALSE}
library(psych)
describeBy(df, group) #Returns statistics by group
```
The group argument should be a factor


<hr></hr>

###Functions

Use functions to repeat common actions and to help compartmentalise your code

```{r functions, eval = FALSE}
function_name <- function(argument_1, argument_2, ...){
  do something
  return(return_value)
}
```




</div>
</div>
